% !TEX root = main-arxiv.tex

\bigskip\noindent\textbf{** Begin Developer Notes **}

% \bartnote{Decided to focus the paper on HDSA with respect to model discrepancy. We can mention in the intro and/or conclusion that there are other outer-loop problems that we can use the \SABL $\to$ \MrHyDE approach for, but let the main focus be on HDSA throughout.}

\bartnote{The big thing about \HdsaLib is that it implements the model discrepancy details, abstracts the interface and requires the user to instantiate the things that are needed. The high-level challenge is accessibility because of the complexity of the HPC environment. Need to point to the abstractions, i.e., make sure any sample code listings are tied carefully to the mathematics.}

\shanenote{Explaining the full OOP structure of \HdsaLib would be too much, but showing an example of a particular piece of the algorithm in all three elements might be effective. Need to drive everything from a very high-level view. Maybe have a baby 1D problem just in \SABL (everything is plottable but it doesn't take much space), then a transition problem done in both \SABL and \MrHyDE, then a monster problem done solely in \MrHyDE. Can make the case that this is one way to validate your HPC implementations as well as lower the barrier to entry for new HPC practitioners.}

\bartnote{Implementation strategy is really step by step, to the level of verifying matvecs at each stage of the algorithm. Also, \MrHyDE is a particular driver, but not the only driver for \HdsaLib. Maybe have Section~3 be \HdsaLib, Section~4 HPC simulator with \MrHyDE as an example.}

\shanenote{Give the sense that this is very complicated. Don't underplay the complexity of the full-order model (finite element is already hard, it could be a time-dependent problem), solving the optimization problem (trust region methods, Newton Krylov, etc.), the HDSA-MD Bayesian inference problem (with a prior, etc.), the posterior update, sampling, etc. The high-level accessibility is still very important, we want it to be publishable but also readable.}

\bartnote{Do we have to show the final product of the $\H$ and $\B$ and the huge inverse to get $\bfSigma$? Should think about if we want to show any more detail. Might want to just list the big expressions to better show what the ingredients at the expense of detailed derivation? Could be philosophical up front, then walk through the pieces. Forward modeling, Bayesian modeling and sampling, etc. \SABL could be a big section, walk through an example slowly? This paper should be \textbf{focused on the development paradigm}, not feel like a practitioner-oriented tutorial paper.}

\bartnote{Diagram for under-the-hood linear algebra? Might even re-purpose the diagram from the hyperparameter sampling paper. Let's try to highlight some of the complexity but not necessarily all of it. Perhaps $\B\bar{\bftheta}$ is sufficient. We can describe in words that the sampling is analytic and based on matrix factorizations, etc. Once we get to \MrHyDE, we probably can't form matrices explicitly and have to do something special to do matrix-free computations. Might want a table that mirrors \Cref{table:hdsa-requirements} that lists math on the left, then shows the tools needed in \MrHyDE for each thing. Don't have to make too strong of a distinction between the frontend and backend pieces. Software design takes place in \SABL, but the algorithmic design happens beforehand in the math (making sure everything can be done efficiently). In \SABL, and for small problems, we don't have to worry yet about the complications of matrix-free operations, sparse data structures, etc. but later we will. That means we can validate the algorithm on small problems without worrying about HPC challenges. Need to get the high-level math to \SABL to \MrHyDE story correct, pick and choose demonstrations that support our points but don't need to be comprehensive in terms of the implementation details.}

\bartnote{For the first numerical example, we want to talk about the workflow and where we switch between \SABL and \HdsaLib for comparing different algorithmic components. In \SABL, check that computations are efficient and that the algorithm works, then verify \HdsaLib/\MrHyDE intermediate steps and final results. Write the formulation, focus on the high-level parts of the story, etc.}

\noindent\textbf{** End Developer Notes **}
\clearpage

\section{Introduction}

\begin{purpose}
High mathematical complexity leads to high implementation difficulty.
This paper is about lowering the difficulty of implementing a class of highly complex problems.
\end{purpose}

\noindent
Mathematics, algorithms, and implementation.
These three steps, sequential but intertwined, form the general approach to problems in computational science and engineering, statistics and data science, and scientific machine learning.
In the first step, a problem is formulated, models are proposed, and mathematical analyses lead to solution strategies; second, the components leading to a solution are grouped, ordered, and articulated as an algorithm; and finally, algorithms are implemented as executable software.
An implementation should then be verified --- checked for correctness --- before validating the overall approach against available test data~\cite{oden2004verification,oden2002estimation}.
Implementation difficulty increases rapidly as the complexity of the mathematical setting and the accompanying algorithms grows.
High-performance computing (HPC) seeks to address the computational challenges that arise from high complexity, but harnessing the power of HPC tools often requires expert-level knowledge and experience.
This paper presents a paradigm for developing algorithms and utilizing HPC tools in the highly complex setting of outer-loop problems.

\begin{purpose}
Outer-loop problems go beyond forward simulation. Introduce and motivate HDSA.
% an outer-loop sensitivity analysis tool that sits on top of / wraps around large-scale constrained optimization problems. Model-form error, etc.
\end{purpose}

In contemporary computational science and engineering, an outer-loop problem is a procedure which requires many expensive ``forward'' simulations of a physical process to enable discovery, control, or decision making.
Examples include inverse problems~\cite{ghattas2021learning,stuart2010bayesianinverse}, optimal experimental design~\cite{alexanderian2021oed,aretz2024greedysensorselection,attia2018oed}, sensitivity analysis~\cite{brandes2007stabilityanalysis,buskens2006parametricsensitivity}, large-scale optimization~\cite{akcelik2006parallelpdecon,biegler2007pdeconstrainedopt,zahr2015progressive}, and uncertainty quantification~\cite{oden2004verification,biehler2014efficientuq,na2012reliability,smith2013uqbook}.
This work focuses on hyper-differential sensitivity analysis~(HDSA) \cite{hart2023hdsaillposed,hart2023hdsaoptimal,hart2025hdsaposterior,hart2020hdsapde,stevens2022hdsaodes,sunseri2024hdsabayesian,sunseri2020hdsainverse}, a recent approach to deriving sensitivities for large-scale optimization problems with partial differential equations (PDEs) as constraints.
In particular, we focus on HDSA in the context of model discrepancy (HDSA-MD), meaning the constraint equations used in the optimization are deficient or only approximate \cite{hart2023hdsaoptimal,hart2025hdsaposterior}.
This is useful for cases in which (1) numerical models that fully and accurately reflect the true PDE dynamics are computationally expensive to solve --- so that performing adjoint-based optimization with such models is infeasible --- but a computationally inexpensive reduced-physics, low-fidelity, or reduced-order numerical model for the governing PDEs is available, or (2) the true system dynamics are unknown, but limited experimental data and an approximate numerical model are available.
The main idea is to solve the optimization problem with the inexpensive-but-deficient model, derive sensitivities of the solution with respect to the model form error, and use those sensitivities to update the solution to better reflect the true dynamics.
This procedure requires many forward and backward/adjoint solves of the inexpensive model, as well as access to a small number of forward solves from the expensive-but-accurate model.
The HDSA-MD framework has been demonstrated to be effective for \SM{@Joey, add short success story statement here?}

\begin{purpose}
HDSA is mathematically complicated and difficult to implement at scale within HPC frameworks.
% HPC has high software complexity, lots of minutia (data types, parallelism and communication, etc.).
\end{purpose}

Like other outer-loop procedures, HDSA-MD is highly complex from mathematical, algorithmic, and implementation viewpoints.
Numerically solving PDEs entails introducing a spatial discretization (e.g., through the finite element method) and often involves complicated meshes, large linear or nonlinear systems, and, in problems with time-dependent dynamics, time integration and stability analysis.
In addition, PDE-constrained optimization requires derivatives of the objective function and constraint equations (leading to backward/adjoint PDE solves), step size selection, intelligent termination criteria, and \SM{other?}.
Furthermore, HDSA-MD \SM{something about sampling from prior distributions, gradient / Hessian applications, etc.}
%
All of these elements are further complicated in HPC-level applications where the problem size necessitates parallelism and custom data structures.
Taken together, this results in a high barrier to entry to new practitioners who desire to deploy or extend HDSA-MD and makes implementing HDSA-MD (and validating that implementation) a challenging prospect.

\begin{purpose}
Goal of this paper: implement HDSA-MD. Do this by writing a high-level library first, then a low-level HPC library.%
% This paper presents two complementary software libraries that aim to lower the barrier to entry for HDSA practitioners. The first is a MATLAB library that implements HDSA in a mathematically intuitive way and provides templates for interfacing with the algorithm. The second is a \cpp library with a similar structure as the \MATLAB library but with state-of-the-art HPC features (parallelism, optimization, etc.).
% \textbf{Goals of this paper}: A minimal-but-sufficient, accessible mathematical presentation of HDSA; present an HPC implementation that abstracts away much of the complexity inherent in HDSA and HPC.
\end{purpose}

This paper addresses these challenges by presenting a software solution for HDSA-MD that, to the extent possible, abstracts the mathematical complexity and defines accessible user interfaces.
In doing so, we advocate for a two-step software development strategy: first, write a library in a high-level language (\MATLAB, \texttt{Python}, etc.) to express algorithms in an intuitive and mathematical form, make rapid design choices and refinements, and validate the method on simple problems; second, use the high-level library as a guide to develop an HPC library in a low-level, high-performance language (\cpp, \texttt{FORTRAN}, etc.).
The first library can also be used to some extent to verify the second library.
We demonstrate this paradigm by presenting \MATLAB and \cpp libraries that implement HDSA-MD.
% The high-level library is used to validate the low-level library on small test problems, then the low-level library is used to tackle problems at scale.
Our objectives are twofold: (1) demonstrate a principled approach to implementing complex mathematical frameworks, and (2) lower the barrier to entry for using and/or extend HDSA-MD.
Although this paper focuses on HDSA-MD, the high-to-low-level paradigm may used as an effective implementation strategy for other outer-loop algorithms.

\begin{purpose}
Outline the paper.
\end{purpose}

The remainder of this article is organized as follows.
\Cref{sec:hdsa} outlines HDSA-MD mathematically.
\Cref{sec:sabl} presents \SABL, a (high-level) \MATLAB library that implements PDE-constrained optimization and HDSA-MD.
% The library is used to solve a prototypical problem.
This library is used as a prototype for the (low-level) \cpp library \HdsaLib, described in \Cref{sec:hpc}, which implements HDSA-MD with high-performance numerical tools.
% We also show how to as well as \MrHyDE, a \cpp library with powerful forward solver capabilities.
\Cref{sec:application} presents a large-scale HPC problem that is solved with \HdsaLib,
% and \MrHyDE,
and \Cref{sec:conclusion} offers some concluding remarks.
% Two representative demonstrations: a simple problem (probably advection-diffusion) implemented in \SABL, then \MrHyDE; and a large-scale problem requiring the HPC tools available within \MrHyDE.

\section{Hyper-differential sensitivity analysis with respect to model discrepancy}
\label{sec:hdsa}

This section gives a mathematical description of HDSA-MD, which leverages post-optimality sensitivities in the presence of model form insufficiencies to solve PDE-constrained optimization problems with expensive or inaccessible constraint models.
\Cref{sec:optimization} establishes the setting of large-scale constrained optimization problems with adjoint-based solution strategies.
The problem is modified to account for discrepancies in the model form in \Cref{sec:model-discrepancy}, and \Cref{sec:hdsa-algorithm} outlines a strategy for using limited high-fidelity data to improve the quality of the solution despite model inaccuracies.

\subsection{PDE-constrained optimization}
\label{sec:optimization}

Consider the following constrained minimization problem:
\begin{align}
    \label{eq:optimization-constrained}
    \begin{aligned}
    \min_{\u,\z} ~& J(\u,\z)
    \qquad
    \text{where}~~& \c(\u,\z) = \0.
    \end{aligned}
\end{align}
Here, $\u\in\R^{n_u}$ is the state variable, $\z\in\R^{n_z}$ is the control variable, $J:\R^{n_u}\times\R^{n_z}\to\R$ is the objective function, and $\c:\R^{n_u}\times\R^{n_z}\to\R^{n_u}$ specifies the constraints.
% Here, $n_u\in\mathbb{N}$ is the state dimension and $n_z\in\mathbb{N}$ is the control dimension.
In the context of PDE-constrained optimization, $\c$ represents the residual of a fully discretized PDE, for example after applying a finite element discretization and, if the problem is time-dependent, a time integration scheme.
If $\c(\cdot,\z)=\0$ admits a unique state solution over an admissible set of controls $Z \subset\R^{n_z}$ --- that is, if the discretized PDE is well posed for $\z\in Z$ --- then there exists a solution operator $\s:Z\to U \coloneqq \s(Z) \subset \R^{n_u}$ such that $\c(\s(\z),\z)=\0$ for all $\z\in Z$.
In this case, \cref{eq:optimization-constrained} can be formulated as the equivalent optimization problem
\begin{align}
    \label{eq:optimization-reducedspace}
    \min_{\z\in Z} \hat{J}(\z)
    \coloneqq J(\s(\z),\z).
\end{align}
The problem~\cref{eq:optimization-reducedspace} is often called the \emph{reduced-space formulation} of \cref{eq:optimization-constrained}.

Local minima of \cref{eq:optimization-reducedspace} can be computed with standard optimization strategies as long as the derivatives of $\hat{J}$ can be computed efficiently.
% gradient $\grad{\z}\hat{J}(\z)\in\R^{n_z}$ and, for any vector $\v\in\R^{n_z}$, Hessian-vector products $\v\mapsto \grad{z,z}\hat{J}(\z) \v \in \R^{n_z}$ can be computed efficiently.
% {prf:ref}`alg:adjoint_gradient` shows how to efficiently calculate the gradient by utilizing adjoint-based derivative formulas.
% Similarly, {prf:ref}`alg:adjoint_hessvec` shows how to compute Hessian-vector products using incremental state and incremental adjoint equations, which avoids explicitly forming the (very large) Hessian matrix $\grad{z,z}\hat{J}(\z)\in\R^{n_z \times n_z}$.
The adjoint method~\cite{courant1962physics,givoli2021adjoint,tarantola2005inverse} provides expressions for the derivatives of $\hat{J}$ in terms of the (partial) derivatives of $J$ and $\c$.
Applying the chain rule,
\begin{align}
    \label{eq:gradJ-chain-rule}
    \grad{z}\hat{J}(\z)
    = \nabla[J(\s(\z),\z)]
    = \s_z(\z)\trp\grad{u}J(\s(\z),\z) + \grad{z}J(\s(\z),\z),
\end{align}
where $\s_z(\z)\in\R^{n_u\times n_z}$ is the $\z$ Jacobian of $\s$.
Note that we are using the convention that the gradient is a column vector.\footnote{If the gradient is defined as a row vector, the mathematics are equivalent but read slightly differently. For example,~\cref{eq:gradJ-chain-rule} would be $\grad{z}\hat{J}(\z) = \grad{u}J(\s(\z),\z)\s_z(\z) + \grad{z}J(\s(\z),\z)$. The choice of convention does not affect the implementation.}
Since $\c(\s(\z),\z) = \0$, and assuming that $\c$ and $\s$ are sufficiently smooth, the implicit function theorem yields the formula
\begin{align}
    \label{eq:Sz-implicit-function}
    \s_z(\z)
    = -\c_u(\s(\z),\z)^{-1}\c_z(\s(\z),\z),
\end{align}
where $\c_u(\u,\z)\in\mathbb{R}^{n_u\times n_u}$ and $\c_z(\u,\z)\in\mathbb{R}^{n_u\times n_z}$ denote the $\u$ and $\z$ Jacobians of $\c$, respectively.
Writing $\u = \s(\z)$, the first term in \cref{eq:gradJ-chain-rule} becomes
\begin{align}
    \begin{aligned}
    \s_z(\z)\trp\grad{u}J(\s(\z),\z)
    &= \left(-\c_u(\u,\z)^{-1}\c_z(\u,\z)\right)\trp\grad{u}J(\u,\z)
    \\
    &= -\c_z(\u,\z)\trp\c_u(\u,\z)\invtrp\grad{u}J(\u,\z)
    = \c_z(\u,\z)\trp\bflambda,
    \end{aligned}
\end{align}
where $\bflambda \coloneqq -\c_u(\u,\z)\invtrp\grad{u}J(\u,\z) \in \R^{n_u}$ is the \emph{adjoint state} and resides in the dual of the state space.
Hence, for a given $\z$, the gradient $\grad{z}\hat{J}(\z)$ in \cref{eq:gradJ-chain-rule} can be computed by assembling or applying the derivatives of $J$ and $\c$, as summarized in \Cref{alg:adjoint-gradient}.
Similar arguments lead to an expression for the product of the Hessian $\grad{z,z}\hat{J}(\z) \in\R^{n_z\times n_z}$ with a vector $\v\in\R^{n_z}$ in terms of the second-order derivatives of $J$ and $\c$, detailed in \Cref{alg:adjoint-hessvec}.
With these derivatives on hand, \cref{eq:optimization-reducedspace} can be solved with a Newton--Raphson approach, i.e., a variation on the iteration
\begin{align}
    \label{eq:newton-cg}
    \z_{k+1}
    = \z_{k} - \gamma_k\boldsymbol{\eta}_k,
    \qquad
    \grad{z,z}\hat{J}(\z_{k})\boldsymbol{\eta}_{k} = \grad{z}\hat{J}(\z_{k}),
\end{align}
where $\gamma_k > 0$ is the step size, chosen to satisfy standard Wolfe or Armijo conditions~\cite{armijo1966minimization,wolfe1969conditions}.
The linear system for the step direction $\boldsymbol{\eta}_{k}$ may be solved, for example, via the conjugate gradient method~\cite{hestenes1952conjugategradient}, which only requires Hessian-vector applications $\v\mapsto\grad{z,z}\hat{J}(\z_k)\v$ instead of forming $\grad{z,z}\hat{J}(\z_k)$ explicitly.
Note that the computational cost of this approach is typically dominated by the cost of evaluating the solver $\s(\z)$ and solving adjoint equations involving the Jacobians of $\c$.

\begin{algorithm}[t]
\begin{algorithmic}[1]
\Procedure{AdjointGradient}{~$\z\in\R^{n_z}$~}
    % \State $\u \gets \s(\z)\in\R^{n_u}$
    \State Solve $\c(\u, \z) = \0$ for $\u$, i.e., $\u \gets \s(\z)$
        \Comment{Solve the (PDE) constraint.}
    \State Solve $\c_u(\u,\z)\trp \bflambda = - \grad{u}J(\u,\z)$ for $\bflambda$
        \Comment{Compute the adjoint.}
    \State \textbf{return} $\c_z(\u,\z)\trp \bflambda + \grad{z}J(\u,\z)$
        \Comment{Assemble the gradient.}
\EndProcedure
\end{algorithmic}
\caption{Adjoint-based gradient $\grad{z}\hat{J}(\z)$ for minimizing \cref{eq:optimization-reducedspace}.}
\label{alg:adjoint-gradient}
\end{algorithm}

\begin{algorithm}[t]
\begin{algorithmic}[1]
\Procedure{AdjointHessvec}{
    % State
    $\u\in\R^{n_u}$,
    % control
    $\z\in\R^{n_z}$,
    % adjoint
    $\bflambda\in\R^{n_u}$,
    % search direction
    $\v\in\R^{n_z}$
}
    \State Solve $\c_u(\u,\z) \bfmu = -\c_z(\u,\z) \v$ for $\bfmu$
        % \Comment{Compute the incremental adjoint.}
    \State $\x_J \gets \grad{z,u}J(\u,\z) \bfmu +  \grad{z,z}J(\u,\z) \v$
    \State $\y_J \gets \grad{u,u}J(\u,\z) \bfmu + \grad{u,z}J(\u,\z) \v$
    \State $\y_c \gets \bflambda\trp \c_{u,u}(\u,\z) \bfmu +  \bflambda\trp \c_{u,z}(\u,\z) \v$
    \State Solve $\c_u(\u,\z)\trp \bfgamma = -(\y_J + \y_c)$ for $\bfgamma$
    \State $\x_c \gets \c_z(\u,\z)\trp \bfgamma + \bflambda\trp \c_{z,u}(\u,\z) \bfmu + \bflambda\trp \c_{z,z}(\u,\z)\v$
    \State \textbf{return} $\x_J + \x_c$
\EndProcedure
\end{algorithmic}
\caption{Adjoint-based Hessian-vector product $\v\mapsto\grad{z,z}\hat{J}(\z)\v$ for minimizing \cref{eq:optimization-reducedspace}.}
\label{alg:adjoint-hessvec}
\end{algorithm}

% \begin{algorithm}[t]  % Gauss--Newton, probably not relevant here.
% \begin{algorithmic}[1]
% \Procedure{GaussNewtonHessvec}{
%     % State
%     $\u\in\R^{n_u}$,
%     % control
%     $\z\in\R^{n_z}$,
%     % adjoint
%     $\bflambda\in\R^{n_u}$,
%     % search direction
%     $\v\in\R^{n_z}$
% }
%     \State Solve $\c_u(\u,\z) \bfmu = -\c_z(\u,\z) \v$ for $\bfmu$
%         % \Comment{Compute the incremental adjoint.}
%     \State $\x_J \gets \grad{z,u}J(\u,\z) \bfmu +  \grad{z,z}J(\u,\z) \v$
%     \State $\y_J \gets \grad{u,u}J(\u,\z) \bfmu + \grad{u,z}J(\u,\z) \v$
%     \State Solve $\c_u(\u,\z)\trp \bfgamma = -\y_J$ for $\bfgamma$
%     \State $\x_c \gets \c_z(\u,\z)\trp \bfgamma$
%     \State \textbf{return} $\x_J + \x_c$
% \EndProcedure
% \end{algorithmic}
% \caption{Gauss--Newton approximation of the Hessian-vector product $\grad{z,z}\hat{J}(\z)\v $ for minimizing \cref{eq:optimization-reducedspace}.}
% \label{alg:gaussnewton-hessvec}
% \end{algorithm}

A software implementation for solving \cref{eq:optimization-reducedspace} through \cref{eq:newton-cg} with \Cref{alg:adjoint-gradient,alg:adjoint-hessvec} requires several ingredients, namely the constraint solver $\s:\z\mapsto\u$ and the derivatives of $J$ and $\c$.
The software libraries introduced in later sections define abstract templates with methods for each ingredient, to be implemented by the user or handled with automatic differentiation when possible.

\subsection{Model discrepancy sensitivities}
\label{sec:model-discrepancy}

There are several scenarios in which solving \cref{eq:optimization-reducedspace} is computationally infeasible: (1) solving the constraint and/or forming its derivatives is possible, but highly computationally expensive; (2) the constraint is available only as a black box, meaning it can be evaluated but not differentiated; and (3) the constraint is unknown or unavailable for computation.
In each scenario, one remedy
% In large-scale problems, solving the constraint equation $\c(\u, \z) = \0$ or forming the derivatives of $\c$ can be computationally intensive, prohibiting the direct optimization approach described in the previous section. One remedy
is to replace the constraint with a computationally tractable model based on simplified physics, model order reduction, or data-driven surrogate modeling.
To that end, let $\tilde{\c}:\R^{n_u}\times\R^{n_z}\to\R^{n_u}$ be an approximation to the original constraint $\c$ with corresponding solution map $\tilde{\s}:Z\to U$, i.e., $\tilde{\c}(\tilde{\s}(\z), \z) = \0$ for all $\z\in Z$, and consider the reduced-space problem
\begin{align}
    \label{eq:optimization-surrogate}
    % \begin{aligned}
    % \min_{\u,\z} ~& J(\u,\z)
    % \\
    % \text{where}~~& \tilde{\c}(\u,\z) = \0
    % \end{aligned}
    % \qquad\Longleftrightarrow\qquad
    \min_{\z\in Z}
    \tilde{J}(\z) \coloneqq
    J(\tilde{\s}(\z), \z).
\end{align}
The discrepancy between the model constraint $\tilde{\c}$ and the true constraint $\c$ may cause minimizers of \cref{eq:optimization-surrogate} to be far from minimizers of \cref{eq:optimization-reducedspace}.
We will show, however, that solutions of \cref{eq:optimization-reducedspace} can be written in terms of solutions of \cref{eq:optimization-surrogate}.
% in which $\tilde{\c}(\tilde{\s}(\z), \z) = \0$ for all $\z\in Z$.
If $\tilde{\s}$ is less expensive to evaluate than $\s$, and if the derivatives of $\tilde{\c}$ are more easily formed than the derivatives of $\c$, then solving \cref{eq:optimization-surrogate} and strategically updating that solution is a computationally tractable alternative to solving \cref{eq:optimization-reducedspace} directly.

% \SM{Might be good to introduce the baby example here, define $\c$ and $\tilde{\c}$?}

To relate \cref{eq:optimization-surrogate} to \cref{eq:optimization-reducedspace}, let $\bfdelta \coloneqq \bfdelta(\z;\bftheta)\in\R^{n_z}$ for $\bftheta\in\R^{n_\theta}$ be a smooth function, parameterized such that $\bfdelta(\z;\0) = \0$ for all $\z$ (hence $\bfdelta_z(\z;\0)$ is the zero matrix), and consider the perturbed objective function
\begin{align}
    \label{eq:optimization-discrepancy}
    \J(\z, \bftheta)
    \coloneqq J\big(\tilde{\s}(\z) + \bfdelta(\z;\bftheta), \z\big).
\end{align}
We call $\bfdelta(\z,\bftheta)$ the \emph{discrepancy term}.
By construction,
$
    \J(\z,\0)
    % = J(\tilde{\s}(\z) + \bfdelta(\z;\0),\z)
    = J(\tilde{\s}(\z),\z)
    = \tilde{J}(\z),
$
the objective function in the surrogate problem~\cref{eq:optimization-surrogate}.
On the other hand, if $\bftheta$ is such that $\tilde{\s}(\z) + \bfdelta(\z;\bftheta) \approx \s(\z)$ for all $\z$, then
$
    \J(\z, \bftheta)
    % = J(\tilde{\s}(\z) + \bfdelta(\z;\bftheta), \z)
    \approx
    J(\s(\z), \z)
    = \hat{J}(\z),
$
the objective function for the original optimization problem \cref{eq:optimization-reducedspace}.

Next, let $\tilde{\z}\in\R^{n_z}$ be a local minimizer of~\cref{eq:optimization-surrogate}.
Fixing $\bftheta = \0$, the necessary optimality conditions give
\begin{align*}
    \grad{z}\J(\z,\0)
    = \grad{z}\tilde{J}(\z)
    = \0
\end{align*}
and that $\grad{z,z}\J(\z,\0)$ is positive definite.
By the implicit function theorem, there exists a smooth function $\f:\R^{n_\theta}\to\R^{n_z}$ such that $\f(\0) = \tilde{\z}$ and $\grad{z}\J(\f(\bftheta), \bftheta) = \0$ for all $\bftheta$ in a neighborhood of $\bftheta = \0$, where $\f$ has the $n_z \times n_\theta$ Jacobian
\begin{align}
    \label{eq:theta-jacobian}
    \f_\theta(\bftheta)
    = - \grad{z,z}\J(\f(\bftheta), \bftheta)^{-1}\grad{z,\theta}\J(\f(\bftheta), \bftheta).
    % = -\H(\bftheta)^{-1}\B(\bftheta),
\end{align}
Now let $\z'$ be a solution to the original problem \cref{eq:optimization-reducedspace}.
If we can find $\bftheta'$ such that $\z' = \f(\bftheta')$, then linearizing $\f$ around $\bftheta = \0$ yields the approximation
\begin{align}
    \label{eq:control-update}
    \z'
    = \f(\bftheta')
    \approx \f(\0) + \f_\theta(\0)\bftheta'
    = \tilde{\z} + \f_\theta(\0)\bftheta',
\end{align}
which is an update formula from the solution $\tilde{\z}$ of \cref{eq:optimization-surrogate} to the solution $\z'$ of \cref{eq:optimization-reducedspace}.

\subsection{Hyper-differential sensitivity analysis}
\label{sec:hdsa-algorithm}

Computing $\z'$ with \cref{eq:control-update} requires three components:
the surrogate problem solution $\tilde{\z}$, which can computed by applying the adjoint method to \cref{eq:optimization-surrogate};
the Jacobian $\f_\theta$ given in \cref{eq:theta-jacobian} at $\bftheta=\0$;
and the optimal discrepancy parameters $\bftheta'$, to be chosen so that $\tilde{\s}(\z) + \bfdelta(\z;\bftheta') \approx \s(\z)$.
\Cref{alg:adjoint-gradient,alg:adjoint-hessvec} describe the ingredients needed to compute $\tilde{\z}$ (replacing $\c$ with $\tilde{\c}$, $\s$ with $\tilde{\s}$, and $\hat{J}$ with $\tilde{J}$).
For $\f_\theta(\0)$, let $\H \coloneqq \grad{z,z}\J(\f(\0),\0) \in \R^{n_z\times n_z}$ and note that, since $\bftheta=\0$, $\H = \grad{z,z}\tilde{J}(\tilde{\z})$, which is calculated in \Cref{alg:adjoint-hessvec}.
The remaining piece is
$\B \coloneqq \grad{z,\theta}\J(\f(\0),\0)\in\R^{n_z\times n_\theta}$ so that $\f_\theta(\0) = -\H^{-1}\B$.
Observe, with some abbreviation, that
\begin{align}
    \begin{aligned}
    \label{eq:B-expanded}
    \grad{z,\theta}\J(\z, \bftheta)
    % &= \grad{\theta}\grad{z}[J(\tilde{\s}(\z) + \bfdelta(\z;\bftheta),\z)]
    &= \grad{\theta}\grad{z}[J(\tilde{\s} + \bfdelta,\z)]
    % \\
    % &= \grad{\theta}[
    %     (\tilde{\s}_z(\z) + \bfdelta_z(\z;\bftheta))\trp\grad{u}J(\tilde{\s}(\z) + \bfdelta(\z;\bftheta),\z)
    %     + \grad{z}J(\tilde{s}(\z)+\bfdelta(\z;\bftheta), \z)
    % ]
    \\
    &= \grad{\theta}[
        (\tilde{\s}_z + \bfdelta_z)\trp\grad{u}J(\tilde{\s} + \bfdelta,\z)
        + \grad{z}J(\tilde{\s} + \bfdelta, \z)
    ]
    \\
    &= (\tilde{\s}_{z} + \bfdelta_{z})\trp\grad{u,u}J(\tilde{\s} + \bfdelta, \z)\bfdelta_\theta
    + \grad{u}J(\tilde{\s} + \bfdelta, \z)\trp\bfdelta_{z,\theta}
    + \grad{z,u}J(\tilde{\s} + \bfdelta, \z)\bfdelta_\theta.
    \end{aligned}
\end{align}
Evaluated at $(\z,\bftheta) = (\f(\0),\0) = (\tilde{\z},\0)$ and recalling $\bfdelta_z(\z,\0) = \0$, this simplifies to give
\begin{align}
    \label{eq:B-matrix}
    \B
    % = \tilde{\s}_z(\tilde{\z})\trp\grad{u,u}J(\tilde{\s}(\tilde{\z}),\tilde{\z})\bfdelta_\theta(\0)
    % + \grad{u}J(\tilde{\s}(\tilde{\z}),\tilde{\z})\trp\bfdelta_{z,\theta}(\0)
    % + \nabla_{z,u}J(\tilde{\s}(\tilde{\z}), \tilde{\z})\bfdelta_\theta(\0).
    = \tilde{\s}_z(\tilde{\z})\trp\grad{u,u}J(\tilde{\u},\tilde{\z})\bfdelta_\theta(\tilde{\z},\0)
    + \grad{u}J(\tilde{\u},\tilde{\z})\trp\bfdelta_{z,\theta}(\tilde{\z},\0)
    + \nabla_{z,u}J(\tilde{\u}, \tilde{\z})\bfdelta_\theta(\tilde{\z},\0),
\end{align}
where $\tilde{\u} \coloneqq \tilde{\s}(\tilde{\z})$.
%
Because $\H$ is independent of the discrepancy term $\bfdelta$, and since $\B$ depends only on $\bfdelta_\theta(\tilde{\z},\0)$ and $\bfdelta_{z,\theta}(\tilde{\z},\0)$, an application of the Riesz representation theorem leads to a linear-affine form for $\bfdelta$,
\begin{align}
    \label{eq:affine-discrepancy}
    \bfdelta(\z,\bftheta) = \left[\begin{array}{cc}
        \I_{n_u} & \I_{n_u}\otimes\z\trp\M_z
    \end{array}\right]\bftheta,
\end{align}
where $\otimes$ denotes the Kronecker product~\cite{vanloan2000kronecker} and $\M_z\in\R^{n_z\times n_z}$ is a mass matrix defining an appropriate inner product $\langle\z_1,\z_2\rangle = \sqrt{\z_1\trp\M_z\z_2}$ on $Z\subset\R^{n_z}$, the space of admissible controls (see \cite{hart2023hdsaoptimal} for more detail).
Hence, the dimension of $\bftheta$ is $n_\theta = n_u(n_z + 1)$.
This Kronecker structure of \cref{eq:affine-discrepancy} enables efficient matrix-vector products $\v\mapsto\B\v$ without assembling $\B$ explicitly, which we will discuss shortly.

% \begin{purpose}
%     Calibrating $\bftheta$ from data.
%     A prior for $\bftheta$ can be written in terms of priors for $\u$ and $\z$.
%     % Don't go too far into the weeds (e.g., Kronecker structure), just list the essential ingredients that users of \HdsaLib will need to know about in order to use it.
%     % Concise summary, consider a diagram like Figure~1 from \cite{hart2023hdsaoptimal}.
% \end{purpose}

The final item needed for \cref{eq:control-update} is an optimal $\bftheta'$ such that $\bfdelta(\z;\bftheta') \approx \s(\z) - \tilde{\s}(\z)$.
Although solving~\cref{eq:optimization-reducedspace} directly is intractable due to the derivatives of the full constraint $\c$ being inaccessible or highly expensive, we assume that we can either (1) solve the constraint equation (i.e., evaluate $\s(\z)$) a limited number of times or (2) in a data-driven scenario, observe a limited number of state solutions with known controls.
To that end, for a linearly independent set of controls $\{\z_{\ell}\}_{\ell=1}^{n_d}$ for which we have access to $\s(\z_1),\ldots,\s(\z_{n_d})$ and $\tilde{\s}(\z_1),\ldots,\tilde{\s}(\z_{n_d})$, we consider the Bayesian linear inference problem $\A\bftheta = \mathbf{d} + \bfeta$, where
\begin{align}
    \label{eq:bayesian-inference}
    \A =
    \left[\begin{array}{cc}
        \I_{n_u} & \I_{n_u}\otimes\z_1\trp\M_z\\
        \vdots & \vdots \\
        \I_{n_u} & \I_{n_u}\otimes\z_{n_d}\trp\M_z
    \end{array}\right],
    % \in\mathbb{R}^{n_u n_d \times n_\theta},
    \quad
    \mathbf{d}
    = \left[\begin{array}{c}
        \s(\z_1) - \tilde{\s}(\z_1)
        \\ \vdots \\
        \s(\z_{n_d}) - \tilde{\s}(\z_{n_d})
    \end{array}\right],
    % \in\mathbb{R}^{n_u n_d},
    \quad
    \bfeta
    = \left[\begin{array}{c}
        \bfepsilon_1
        \\ \vdots \\
        \bfepsilon_{n_d}
    \end{array}\right],
    % \in\mathbb{R}^{n_u n_d},
\end{align}
and in which each $\bfepsilon_{\ell}\sim\mathcal{N}(\0, \alpha\M_u^{-1})$ is an uncorrelated Gaussian noise term with variance scaling $\alpha > 0$ and where $\M_u\in\R^{n_u\times n_u}$ is a mass matrix for the state space.
% Note that the $n_u n_d \times n_u(1 + n_z)$ data matrix $\A$ is large and that the system is likely underdetermined.
To develop a prior distribution for $\bftheta$, we impose priors for the state and control spaces.
Let $\W_u\in\R^{n_u\times n_u}$ and $\W_z\in\R^{n_z\times n_z}$ be precision (inverse covariance) matrices for the state and control spaces, respectively.
In the context of PDEs, these precision matrices are often {\color{royalblue}(inverse?)} elliptic operators.
It can be shown that an appropriate precision matrix for a zero-mean prior distribution on the parameter vector $\bftheta$ is
\begin{align}
    \label{eq:prior-parameter-precision}
    \W_\theta
    = \left[\begin{array}{cc}
        \W_u & \W_u \otimes \tilde{\z}\trp\M_z \\
        \W_u\otimes\M_z\tilde{\z} & \W_u\otimes\M_z(\W_z + \tilde{\z}\tilde{\z}\trp\M_z)
    \end{array}\right]
    \in\R^{n_\theta \times n_\theta}.
\end{align}
The posterior distribution for $\bftheta$ is also Gaussian, with the following mean and covariance:
\begin{subequations}
\label{eq:posterior-parameter-distribution}
\begin{align}
    \bar{\bftheta}
    &= \tfrac{1}{\alpha}\bfSigma\A\trp(\I_{n_d}\otimes\M_{u})\mathbf{d},
    \\
    \bfSigma_\theta
    &= \left(\W_{\theta} + \tfrac{1}{\alpha}\A\trp(\I_{n_d}\otimes \M_u)\A\right)^{-1}.
\end{align}
\end{subequations}

Making appropriate choices for the prior distributions, sampling from the posterior distribution \cref{eq:posterior-parameter-distribution}, and propagating the result through the product $\bftheta\mapsto\H^{-1}\B\bftheta$ for \cref{eq:control-update} is nontrivial
% Doing so efficiently involves advanced randomized linear algebra and matrix factorization techniques
and is the subject of several previous works~\cite{hart2023hdsaoptimal,hart2025hdsaposterior}. % TODO: hyperparameter paper.
Computations in the high-dimensional state, control, or parameter spaces can be mitigated by taking advantage of the structure of $\B$ in \cref{eq:B-matrix}.
To make this precise, define
\begin{subequations}
\begin{align}
    \G &= \1\1\trp + (\Z - \tilde{\z}\1\trp)\trp\W_z^{-1}(\Z - \tilde{\z}\1\trp)\in\R^{n_d \times n_d},
    \\
    \Z &= [~\z_1~~\cdots~~\z_{n_d}~]\in\R^{n_z\times n_d},
\end{align}
where $\1\in\R^{n_d}$ is a vector of ones.
Let $(\mu_i,\g_i)$ be the $n_d$ eigenpairs of $\G$, and define
\begin{align}
    \y_i
    &= \Z\g_i - (\1\trp\g_i)\tilde{\z}
    \in\R^{n_z},
    % \\
    % s_i
    % &= (\1\trp\g_i) - \y_i\trp\W_{z}^{-1}\tilde{\z}
    % \in\R,
    \\
    \alpha_{\ell}
    &= 1 - \tilde{\z}\trp\W_{z}^{-1}(\z_\ell - \tilde{\z})
    \in\R,
    \\
    b_{i,\ell}
    &= (\z_\ell - \tilde{\z})\trp\W_{z}^{-1}\Z\g_i + (\1\trp\g_i)\alpha_\ell
    \in \R,
    \\
    \u_{\ell}
    &= \W_u^{-1}\M_u(\s(\z_\ell) - \tilde{\s}(\z_\ell))
    \in\R^{n_u},
    \\
    \u_{\ell,i}
    &= (\alpha \W_u + \mu_i\M_u)^{-1}\M_u \u_\ell
    \in\R^{n_u}
\end{align}
for $i,\ell = 1,\ldots, n_d$.
\end{subequations}
Using the structure of $\B$ in \cref{eq:B-matrix} and of $\bfdelta$ in \cref{eq:affine-discrepancy}, it can be shown \cite{hart2025hdsaposterior} that
% \begin{subequations}
% \begin{align}
%     \label{eq:theta-decomposition}
%     \mathbb{E}[\bftheta]
%     = \overline{\bftheta} + \hat{\bftheta} + \breve{\bftheta}
% \end{align}
% such that
\begin{align}
    \label{eq:theta-propagation}
    \begin{aligned}
    \B\bar{\bftheta}
    &= \frac{1}{\alpha}\tilde{\s}_{z}(\tilde{\z})\trp\grad{u,u}J(\tilde{\u},\tilde{\z})\left(
        \sum_{\ell=1}^{n_d}\left(
            \s(\z_{\ell}) - \sum_{i=1}^{n_d} b_{i,\ell}(\1\trp\g_i)\u_{i,\ell}
        \right)
    \right)
    \\ &\qquad
    + \frac{1}{\alpha}\sum_{\ell=1}^{n_d}
        (\grad{u}J(\tilde{\u},\tilde{\z})\trp\u_{\ell})\W_{z}^{-1}(\z_\ell - \tilde{\z})
    % \\ &\qquad
    + \frac{1}{\alpha}\sum_{\ell=1}^{n_d}\sum_{i=1}^{n_d}b_{i,\ell}(\grad{u}J(\tilde{\u},\tilde{\z})\trp\u_{i,\ell})\W_{z}^{-1}\y_i.
    \end{aligned}
    % \\
    % \B\hat{\bftheta}
    % &= \sqrt{\alpha}\nabla_{z}\tilde{\s}_{z}(\tilde{\z})\trp\grad{u,u}J(\tilde{\u},\tilde{\z})\left(\sum_{i=1}^{n_d}\frac{\1\trp\g_i}{\sqrt{\mu_i}}\hat{\u}_i\right)
    % + \sqrt{\alpha}\sum_{i=1}^{n_d}\frac{\grad{u}J(\tilde{\u},\tilde{\z})\trp\hat{\u}_{i}}{\sqrt{\mu_i}}\W_{z}^{-1}\y_i,
    % \\
    % \B\breve{\bftheta}
    % &= \sum_{k=1}^{n_z - n_d + 1}(\grad{u}J(\tilde{\u},\tilde{\z})\trp\breve{\u}_{k})\W_z^{-1/2}\breve{\z}_k
    % \\
\end{align}
% \end{subequations}
Similar calculations expose structure to enable drawing samples (not just the mean) efficiently from the parameter distribution \cref{eq:posterior-parameter-distribution}.
\SM{Something about how this approach is scalable, no computations in high dimensions, if we didn't take advantage of structure we would be hard pressed to compute anything?}

\begin{table}[t]
\centering
\begin{tabular}{|rl|} %|l}
    % HDSA-MD requirement & \SABL/\HdsaLib class
    % \\ \hline
    \hline
    1. & Solution $(\tilde{\s}(\tilde{\z}),\tilde{\z})$ to surrogate problem \cref{eq:optimization-surrogate}
    \\
    2. & Data $(\z_{\ell},\tilde{\s}(\z_{\ell}),\s(\z_{\ell}))$ for the full and surrogate constraints, $\ell=1,\ldots,n_d$
    % & \texttt{MD\_Data\_Interface}
    \\ % \hline
    3. & Application of surrogate constraint Jacobian transpose, $\v\mapsto\tilde{\s}_{z}(\z)\trp\v$ % Apply_Solution_Operator_z_Jacobian_Transpose
    \\
    4. & Application of surrogate reduced-space Hessian, $\v\mapsto\tilde{J}_{z,z}(\tilde{\z})\v$ % Apply_RS_Hessian
    \\
    5. & Objective state gradient, $\grad{u}J(\u,\z)$ % Misfit_Gradient
    \\
    6. & Application of objective state Hessian, $\v\mapsto\grad{u,u}J(\u,\z)\v$ % Apply_Misfit_Hessian
    % & \texttt{MD\_Opt\_Prob\_Interface}
    \\ % \hline
    7. & Application of state mass matrix, $\v\mapsto\M_u\v$ % Apply_M_u
    \\
    8. & Application of state covariance matrix, $\v\mapsto\W_u^{-1}\v$ % Apply_W_u_Inverse
    \\
    9. & Application of inverse for $\u_{\ell,i}$, $\v\mapsto(\W_u + \alpha\M_u)^{-1}\v$ % Apply_W_u_Plus_scalar_M_u_Inverse
    % Sample_with_Covariance_W_u_Inverse
    % Sample_with_Covariance_W_u_Plus_scalar_M_u_Inverse
    % & \texttt{MD\_u\_Prior\_Interface}
    \\ % \hline
    10. & Application of control mass matrix, $\v\mapsto\M_z\v$ % Apply_M_z
    \\
    11. & Application of control covariance matrix, $\v\mapsto\W_z^{-1}\v$ % Apply_W_z_Inverse
    % Sample_with_Covariance_W_z_Inverse
    % Apply_W_z
    % & \texttt{MD\_z\_Prior\_Interface}
    \\ \hline
\end{tabular}
\caption{Problem-specific requirements for HDSA-MD, designed as matrix-free operations.}
\label{table:hdsa-requirements}
\end{table}

A diagram of HDSA-MD is provided in Figure~TODO.
Fully specifying an HDSA-MD problem requires the solution to the surrogate optimization problem \cref{eq:optimization-surrogate}; data for the Bayesian inference~\cref{eq:bayesian-inference}; derivatives of $J$, $\tilde{J}$, and $\tilde{\s}$; mass and precision matrices for the state and control spaces; and an inverse of the form $(\alpha\W_u + \mu_i\M_u)^{-1}$ for constructing $\u_{\ell,i}$.
These requirements are listed in \Cref{table:hdsa-requirements}.
Our goal now is to design a software framework for implementing this complex procedure.

\section{\SABL: Prototyping HDSA-MD in \MATLAB}
\label{sec:sabl}

% \begin{purpose}
% Why \SABL exists. Instead of throwing newcomers headlong into \cpp, it's better to start in a high-level language.
% \end{purpose}

Implementing HDSA-MD poses two major challenges.
First, the mathematics are sufficiently complicated that a successful implementation necessitates careful verification at each step of the algorithm.
Second, the kinds of problems for which HDSA-MD is most useful are large scale, likely requiring HPC resources and distributed computing.
These are competing and conflicting interests that influence our choice of software language: on the one hand, a high-level language --- designed for high readability and ease of use --- would enable an intuitive translation of mathematics to code and simplify the process of checking intermediate steps, but may ultimately struggle to perform at scale; on the other hand, a low-level language --- optimized for performance but not for readability or implementation ease --- would require substantial expertise and be difficult to verify, but a successful implementation would ultimately be able to tackle large-scale problems.
To address the first issue, this section presents \SABL: \textbf{S}andbox for \textbf{A}adjoint-\textbf{B}ased outer \textbf{L}oop analysis, a \MATLAB library that implements PDE-constrained optimization and HDSA-MD.
The source code is publicly available at \textbf{TODO}.
Our purpose in this discussion is not to provide detailed package documentation, but rather to shed light on the design and implementation process and the associated challenges.
We then use \SABL as a guide for developing a low-level HPC implementation of HDSA-MD in \cpp, which is the topic of \Cref{sec:hpc}.

% \begin{purpose}
% Why \MATLAB: intuitive/mathematical high-level syntax, strong backward compatibility, low barrier to entry, takes care of low-level computing tasks in the background, common/popular language in current engineering curriculum.
% Also, \SABL is Designed in such a way to emulate a \cpp HPC framework: object-oriented design, use \texttt{this} instead of \texttt{obj}, etc.
% \end{purpose}

\begin{remark}[Why \MATLAB?]
Other popular high-level languages such as \texttt{Python} and \texttt{Julia} could also be suitable for a prototype library, but \MATLAB has a few features that make it particularly attractive for our purposes.
It has a low barrier to entry, intuitive syntax, and is designed around linear algebra operations, which form the majority of the mathematics in HDSA-MD.
% At the same time, \MATLAB is interpreted (not compiled) and does not require imports or type declarations, all of which generally lower implementation cost compared to compiled languages.
Furthermore, \MATLAB features an object-oriented paradigm that closely mirrors \cpp class syntax, including \texttt{private} / \texttt{public} blocks and abstract methods (virtual functions).
Unlike open-source languages, \MATLAB also has strong backwards compatibility guarantees, leading to greater code longevity and no recurring maintenance needs.
\end{remark}

\subsection{Abstract interface design}

% \begin{purpose}
% Isolate the abstract information that the user must define.
% \end{purpose}

\noindent
\SABL is written with the object-oriented \MATLAB protocol, which translates relatively easily to \cpp.
We begin by defining template classes with abstract methods (pure virtual functions) that represent the mathematical ingredients needed for HDSA-MD.
To solve a particular problem, the user defines child classes to implement the abstract methods, then uses additional library tools to combine the pieces and execute the algorithm.

\begin{listing}[t]
\begin{minted}{matlab}
classdef Objective < handle

  methods (Abstract, Access = public)  % Objective function derivatives.
    [val, grad_u, grad_z] = J(this, u, z)   % $J(\u,\z)$, $\grad{u}J(\u,\z)$, and $\grad{z}J(\u,\z)$
    [u_out] = J_uu_Apply(this, u_in, u, z)  % $\v\mapsto\grad{u,u}J(\u,\z)\v$
    [u_out] = J_uz_Apply(this, z_in, u, z)  % $\v\mapsto\grad{u,z}J(\u,\z)\v$
    [z_out] = J_zu_Apply(this, u_in, u, z)  % $\v\mapsto\grad{z,u}J(\u,\z)\v$
    [z_out] = J_zz_Apply(this, z_in, u, z)  % $\v\mapsto\grad{z,z}J(\u,\z)\v$
  end

  methods (Access = public)            % Finite difference checks.
    function Finite_Difference_Gradient_Check(this, u, z)
      % Do finite differences of $J$ to check $\grad{u}J$ and $\grad{z}J$
    end
    function Finite_Difference_Hessian_Check(this, u, z)
      % Do finite differences of $\grad{u}J$ to check $\grad{u,u}J$, etc.
    end
  end
end
\end{minted}
\caption{Outline of the \MATLAB class for full-space objective functions $J(\u,\z)$ in \SABL.}
\label{code:sabl-optimization}
\end{listing}

% % Potential additional listings, but probably overkill.
% classdef Constraint < handle
%   methods (Abstract, Access = public)
%     [u] = State_Solve(this, z)            % Solve $\c(\u,\z)$ for $\u$, i.e., evaluate $\s(\z)$
%     [u_out] = c_u_Transpose_Inverse_Apply(this, u_in, u, z)  % $\v\mapsto\c_u(\u,\z)^{-\mathsf{T}}\v$
%     [z_out] = c_z_Transpose_Apply(this, u_in, u, z)          % $\v\mapsto\c_z(\u,\z)\trp\v$
%     [u_out] = c_u_Inverse_Apply(this, u_in, u, z)            % $\v\mapsto\c_u(\u,\z)^{-1}\v$
%     [u_out] = c_z_Apply(this, z_in, u, z)                    % $\v\mapsto\c_z(\u,\z)\v$
%     [u_out] = c_uu_Apply(this, u_in, u, z, lambda)           % $(\v,\bflambda) \mapsto \bflambda\trp\c_{u,u}(\u,\z)\v$
%     [u_out] = c_uz_Apply(this, z_in, u, z, lambda)           % $(\v,\bflambda) \mapsto \bflambda\trp\c_{u,z}(\u,\z)\v$
%     [z_out] = c_zu_Apply(this, u_in, u, z, lambda)           % $(\v,\bflambda) \mapsto \bflambda\trp\c_{z,u}(\u,\z)\v$
%     [z_out] = c_zz_Apply(this, z_in, u, z, lambda)           % $(\v,\bflambda) \mapsto \bflambda\trp\c_{z,z}(\u,\z)\v$
%   end
% end

% classdef Reduced_Space_Optimization < handle
%   methods (Access = public)
%     ReducedSpaceOptimization(objective, constraint)
%     [u, z] = Optimize(this, z0)
%     [val, grad, hessian_data] = Jhat(this, z)         % $\hat{J}(\z)$, $\grad{z}\hat{J}(\z)$
%     [z_out] = Jhat_hessVec(this, hessian_data, z_in)  % $\v\mapsto\grad{z,z}\hat{J}(\z)\v$
%   end
% end

First, we implement \Cref{alg:adjoint-gradient,alg:adjoint-hessvec} for adjoint-based optimization.
Template classes are defined to represent the objective $J$ (or $\tilde{J}$) and the constraint $\c$ (or $\tilde{\c}$).
To verify the implementation of the derivatives of $J$ and $\c$, we include finite difference checks that compare, for example, finite differences of evaluations of $J$ with its state gradient,
\begin{align}
    \frac{1}{\delta}(J(\u + \delta \e_i, \z) - J(\u))
    \approx \e_i\trp \grad{u}J(\u, \z) = (\grad{u}J(\u,\z))_{i},
\end{align}
where $\e_i$ is the $i$-th canonical basis vector, which is checked for various small $\delta$.
\Cref{code:sabl-optimization} shows the outline of the class for the objective.
An additional class is defined that is initialized with instances of the objective and constraint classes, then uses \Cref{alg:adjoint-gradient,alg:adjoint-hessvec} to solve the associated constrained optimization problem via a trust region or Newton--CG iteration.
Taken together, these classes are used to define and solve the surrogate-constrained optimization \cref{eq:optimization-surrogate}, yielding the state-control pair $(\tilde{\s}(\tilde{\z}),\tilde{\z})$.
%  which is the first step for HDSA-MD.

% \SM{I'm learning toward cutting \Cref{code:sabl-optimization} down to just the \texttt{Optimize} class and not naming classes specifically in the body text, and similar for \Cref{code:sabl-hdsa}. The point of this section is to talk about design and the difficulty of actually implementing the backend, so we shouldn't spend a lot space enumerating the API.}

% The user defines child classes that implement the abstract methods; to verify the implementation, each class is provided with methods for comparing user-defined derivatives to finite difference estimates with varying step sizes. Next, the user instantiates the classes and pass them to a \texttt{ReducedSpaceOptimization} class, which solves the corresponding optimization problem via \Cref{alg:adjoint-gradient} and \Cref{alg:adjoint-hessvec} and a Newton--CG method.

\begin{remark}[Interface specialization]
One advantage of an abstract user interface is that it can be specialized to express algorithmic requirements in terms of the mathematical objects that the user is most likely to have access to.
For example, we can extend our interfaces to represent problems in which the state $\u$ represents a time-dependent variable and where the objective and constraints are given by
\begin{gather*}
    \min_{\y(t),\z}~\int_{0}^{T} g(\y(t),t) dt + R(\z)
    \qquad
    \textup{where}
    \qquad
    \ddt\y(t) = \f(\y(t), \z, t), ~~ \y(0) = \h(\z),
\end{gather*}
where
$\y : [0,T] \to \R^{n_y}$,
$g : \R^{n_y} \times [0, T] \to \R$,
$R : \R^{n_z} \to \R$,
$\f : \R^{n_y} \times \R^{n_z} \times [0,T]  \to \R^m$,
and
$\h : \R^{n_z} \to \R^{n_y}$.
In this case, the user must implement the derivatives of $g$, $R$, $\f$, and $\h$, which are then used to assemble the derivatives of the corresponding objective $J$ and constraint $\c$.
% This way, the user does not have to implement $\c$ or its derivatives directly, which involves a time-stepping method.
% Similarly, we define \SM{Elliptic priors?}
\end{remark}

% \begin{remark}[Automatic differentiation]
% Additional templates are available in \SABL that use automatic differentiation via the \texttt{AdiGator} library~\cite{weinstein2017adigator} instead of specifying the derivatives of the objective and constraint explicitly.
% \MATLAB does not natively support automatic differentiation, so...\SM{when it works it works, when it doesn't it's really not great.}
% \end{remark}

\begin{listing}[t]
\begin{minted}{matlab}
classdef MD_u_Prior_Interface < handle
  methods (Abstract, Access = public)
    [u_out] = Apply_M_u(this, u_in)          % $\v\mapsto\M_u\v$
    [u_out] = Apply_W_u_Inverse(this, u_in)  % $\v\mapsto\W_u^{-1}\v$
    [u_out] = Apply_W_u_Plus_scalar_M_u_Inverse(this, u_in, scalar)
  end                                        % $(\v,\alpha)\mapsto(\W_u + \alpha\M_u)^{-1}\v$
end
\end{minted}
\caption{Outline of the \MATLAB class for the state prior in \SABL.}
\label{code:sabl-hdsa}
\end{listing}


% % Potential additional listings, but probably overkill.
% methods (Access = public)            % Methods for other posterior samples.
% function u_out = Sample_with_Covariance_W_u_Inverse(
%   this, num_samples
% )  % Sample from $\mathcal{N}(\0,\W_u^{-1})$
%   error('this method is required for sampling algorithms');
% end
% function u_out = Sample_with_Covariance_W_u_Plus_scalar_M_u_Inverse(
%   this, num_samples, scalar
% )  % Sample from $\mathcal{N}(\0,(\W_u + \alpha\M_u)^{-1})$
%   error('this method is required for sampling algorithms');
% end
% end
%
% classdef MD_z_Prior_Interface < handle
%   methods (Abstract, Access = public)
%     [z_out] = Apply_M_z(this, z_in)           % $\v\mapsto\M_z\v$
%     [z_out] = Apply_W_z_Inverse(this, z_in)   % $\v\mapsto\W_z^{-1}\v$
%   end
% end
%
% classdef MD_Opt_Prob_Interface < handle
%   methods (Abstract, Access = public)                 % $\v\mapsto\tilde{\s}_z(\z)\trp\v$
%     [z_out] = Apply_Solution_Operator_z_Jacobian_Transpose(this, u_in, z)
%     [z_out] = Apply_RS_Hessian(this, z_in, z)         % $\v\mapsto\grad{z,z}\tilde{J}(\z)$
%     [grad_u] = Misfit_Gradient(this, u, z)            % $\grad{u}J(\u, \z)$
%     [u_out] = Apply_Misfit_Hessian(this, u_in, u, z)  % $\v\mapsto\grad{u,u}J(\u, \z)\v$
%   end
% end

Additional problem-specific operations needed to fully specify an HDSA-MD problem include derivatives of $J$, $\tilde{J}$, and $\tilde{\s}$; mass and precision matrices for the state and control spaces; and data for the Bayesian inference~\cref{eq:bayesian-inference}.
\Cref{table:hdsa-requirements} lists these ingredients, posed as matrix-vector products where possible in order to avoid large dense matrix inversions later on.
Requirements 8--11 arise from an efficient sampling of the posterior parameter distribution, which we will discuss shortly.
Similar to the optimization classes, we define template classes with abstract methods for each ingredient, grouped by the surrogate optimization problem (requirements 3--6), the state prior (requirements 7--9), and the control prior (requirements 10--11).
We also include a class for loading the surrogate solution and data (requirements 1--2), which separates solving for $\tilde{\z}$ from the rest of the HDSA-MD procedure.
In an HPC context, this separation means that one code can be responsible for generating data through forward PDE solves and the HDSA-MD update can be computed in a distinct computing environment.

% To enable verification on model problems, \SABL also includes for prototyping and verification, we want to do everything in-house, which means we also write our own forward solvers (constraints).
% But, technically, to do HDSA we just need high-fidelity data and the solution to the surrogate-constrained optimization problem, which could be provided in a separate HPC workflow and imported to an HDSA-MD program later. This is the approach that \HdsaLib takes, but we also include the optimization classes here to get everything in one place.

\subsection{Verification with manufactured solutions}

\SM{FEM verifies with method of manufactured solutions. Comparing code to code is not verification because the first code might be wrong, but we have no other choice. There are a few small things we can do to verify certain things: check for symmetry / positive definiteness, checkpointing quantities throughour algorithmic implementation, etc. We can also try to set up inverse problems for a type of method of manufactured solutions. All of these tasks would be much harder in HPC, but \MATLAB just does it (example: \texttt{eig()} to determine positive definiteness). Maybe focus a bit on how we can verify things (gradients, positive definiteness, symmetry, proptotypical examples to demonstrated improvements in the solution.)
}

\section{\HdsaLib: High-performance HDSA-MD in \cpp}
\label{sec:hpc}

\begin{purpose}
\begin{itemize}
\item Introduce \HdsaLib, scope, repository, etc.
\item High-level comparison with \SABL: more dependencies, lower level, but higher power; defines its own data structures; relies on interfaces to other high-performance codes for the forward solves and differentiation, which we were explicit about in \SABL.
\item Note again that \SABL is designed to prepare users for using \HdsaLib.
\end{itemize}
\end{purpose}

\noindent
Jumping straight to high-performance computing in \cpp with \HdsaLib can be challenging for new users, especially for practitioners who are accustomed to high-level languages such as \MATLAB and Python.

\subsection{Object-oriented design following \SABL}

\begin{purpose}
We've already made design choices, which were easy to quickly change on the fly, while developing \SABL.
Now we simply port it over.
We still have to deal with the low-level nature of \cpp, but we don't have to think so much about the overall algorithmic structure because we've done that part.
\end{purpose}

\begin{listing}[t]
\begin{minted}{cpp}
namespace HDSA {
  template <class RealT> class MD_u_Prior_Interface {
    public:
      virtual void Apply_M_u(
        HDSA::Vector<RealT> & u_out,
        const HDSA::Vector<RealT> & u_in
      ) const = 0;  // $\v\mapsto\M_u\v$

      virtual void Apply_W_u_Inverse(
          HDSA::Vector<RealT> & u_out,
          const HDSA::Vector<RealT> & u_in
      ) const = 0;  // $\v\mapsto\W_u^{-1}\v$

      virtual void Apply_W_u_Plus_scalar_M_u_Inverse(
          HDSA::Vector<RealT> & u_out,
          const HDSA::Vector<RealT> & u_in,
          const RealT & scalar
      ) const = 0;  // $\v\mapsto(\W_u + \alpha\M_u)^{-1}\v$
  };
}
\end{minted}
\caption{Outline of the \cpp class for the prior state distribution in \HdsaLib.}
\end{listing}

% Potential additional listings, probably overkill.
% virtual void Sample_with_Covariance_W_u_Inverse(
%     HDSA::MultiVector<RealT> & samples
%   ) const {}    // Sample from $\mathcal{N}(\0,\W_u^{-1})$

%   virtual void Sample_with_Covariance_W_u_Plus_scalar_M_u_Inverse(
%     HDSA::MultiVector<RealT> & samples,
%     const RealT & scalar
%   ) const {}    // Sample from $\mathcal{N}(\0, (\W_u + \alpha\M_u)^{-1})$

\subsection{\MrHyDE: A high-performance \cpp library for solving differential equations}
\label{sec:mrhyde}

\bartnote{What are the things you need to do for HPC? Derivatives, matrix-vector computations, forward solver, parallelism, etc. In this paper, we're going to use a library that does a lot of things.}

\SM{Note that \MrHyDE takes care of a lot of things for us, like calculating the adjoint at the element level via automatic differentiation. Transposes and handling boundary conditions are nontrivial in a parallel environment, but this is handled by Panzer in \MrHyDE.}

\MrHyDE~also has robust forward solver (inner-loop) capabilities.

% \SM{Not convinced that \HdsaLib should come first. Maybe we should have HDSA first, then \SABL, then a section called ``\HdsaLib and \MrHyDE: implementing HDSA-MD in \cpp''.}

TODO.

\begin{purpose}
\begin{itemize}
\item Introduce \MrHyDE: scope, style, repository, etc.
\item Forward solver capabilities of \MrHyDE.
\item Discuss how \MrHyDE addresses many of the challenges of working in an HPC environment.
\item Emphasize efficient automatic differentiation tools.
\end{itemize}
\end{purpose}

TODO (Tim?).

\subsection{Using \MrHyDE as a driver for \HdsaLib}

TODO

\subsection{Verification by comparison with \SABL}

The complexities of a full HPC environment make it difficult to assess and verify implementations of novel, complicated algorithms such as HDSA-MD.

\section{Example 1: Small problem, done in both \SABL and \HdsaLib}
\label{sec:smallexample}

\begin{purpose}
Define exemplar, something simple (probably advection-diffusion) that allows us to go through the details of \SABL and \HdsaLib.
% Might walk through implementing the Objective and Constraint, and use the ReducedSpaceOptimization to solve the problem.
Show results from each library as validation.
\end{purpose}

\SM{It would be convenient to do something stationary so that we only need to introduce the \texttt{Objective} and \texttt{Constraint} classes instead of also introducing \texttt{Dynamic\_Objective} and \texttt{Dynamic\_Constraint}.}

\section{Example 2: Big fat HPC-worthy problem, \HdsaLib/\MrHyDE only}
\label{sec:application}

\begin{purpose}
Solve a really huge problem that requires parallelism, etc. using \MrHyDE.
Solve a simple version of the problem in \SABL? Probably not.
\end{purpose}

\subsection{Problem statement}

TODO.

\subsection{Implementation considerations}

Maybe?

\subsection{Numerical results}

\section{Conclusion}
\label{sec:conclusion}

TODO.

\begin{purpose}
Summary:
\begin{itemize}
\item We presented two complementary libraries for implementing HDSA, one more intuitive (closer to the mathematical formulae) and the other more powerful. The first is designed to prepare the user for the second.
\item The goal was to be able to implement this complicated algorithm in a mathematically intuitive and high-level context before moving to a low-level but high-performance context.
\item We demonstrated HDSA on a true HPC problem.
\end{itemize}
Lessons learned:
\begin{itemize}
\item This is a really useful paradigm for complicated algorithms.
\MATLAB in particular pairs well with the \cpp libraries we work with.
\end{itemize}
Outlook:
\begin{itemize}
\item Further examples and details in the documentation?
\item By the way, \SABL and \MrHyDE provide solution frameworks for other outer-loop problems, including Bayesian inversion and optimal experimental design (and others?).
\item Ideas for future work?
\end{itemize}
\end{purpose}
